Whose Truth is it Anyway? An Experiment on Annotation Bias in Times of Factual Opinion Polarization

### 2023-07-13 report/draft.Rmd
update draft (conclusion and refs)

---
output: 
  stevetemplates::article:
    fig_caption: true
    extra_dependencies: ["float"]
bibliography: references.bib
urlcolor: black
linkcolor: black
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage[para,online,flushleft]{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{hyperref}
   - \usepackage{array}   
   - \usepackage{caption}
   - \usepackage{graphicx}
   - \usepackage{epstopdf}
   - \usepackage{siunitx}
   - \usepackage{hhline}
   - \usepackage{calc}
   - \usepackage{tabularx}
   - \usepackage{fontawesome}
   - \usepackage{amsthm}
   - \newtheorem{hypo}{Hypothesis}
biblio-style: apsr
title: "Whose Truth is it Anyway? An Experiment on Annotation Bias in Times of Factual Opinion Polarization"
author:
- name: Mariken A.C.G. van der Velden
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
- name: Myrthe Reuver
- affiliation: Computational Linguistics & Text Mining Lab, Vrije Universiteit Amsterdam
- name: Wouter van Atteveldt
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
- name: Antse Fokkens
- affiliation: Computational Linguistics & Text Mining Lab, Vrije Universiteit Amsterdam
- name: Felicia Loecherbach
- affiliation: Center for Social Media & Politics, New York University
- name: Kasper Welbers
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
thanks: "Corresponding author: MACGvdV, Replication files are available on the author's Github account (https://github.com/MarikenvdVelden/bias-experiment); Author contributions: a) designed the study: MACGvdV, MR, WvA, AF, FL, & KW; b) conducted the study: MACGvdV, FL & MR; c) data cleaning & analysis: MACGvdV; d) writing of the paper: MACGvdV, MR, WvA, AF, FL, & KW"
anonymous: FALSE
abstract: "The behavior of citizens and the study of them by social scientists rely heavily on key information. The democratization of data has lead to numerous possibilities for gathering and analyzing textual data. These enourmous amounts of data are typically handled by machine learning techniques to classify into meaningful variables. The performance of these models is evaluated based on a gold standard, created by human annotators. Having a high level of agreement between these annotators is key, but recent research suggests personal characteristics of annotators, like political ideology or knowledge, interfere. We show in two pre-registered experiments that XXX. Thereby [contribution]."
keywords: "Experiment, Annotation Bias, Ideology, Measurinig Political Position, Text-as-Data, Political Knowledge"
geometry: margin=1in
mainfont: cochineal
fontsize: 11pt
params:
  anonymous: ""
doublespacing: TRUE
endnote: no
pandocparas: TRUE
sansitup: FALSE
editor_options: 
  markdown: 
    wrap: sentence
---

@@

# Introduction

The behavior of citizens and the study of them by social scientists rely heavily on key information.
Classical theories of political communication, such as agenda setting or framing [e.g. @vanaelst2016political; @lecheler2019news], formulate that political information drives opinion formation and participation in politics -- from voting to protests [for overviews hereof across countries, see @pfetsch2013comparing].
The democratization of data and advent of computational social science has paved the way for new possibilities of gathering and analyzing textual data [for a recent overview, see @vanatteveldt2022computational]. 
In particular, advances in Natural Language Processing (NLP) have made it possible to automatically analyze large quantities of data using machine learning [e.g., see [ADD REF, Bender does not make sense here] @wei2023overview].
To determine the validity of such large scale computational analyses, we rely on "gold standard" data, created by human annotators.
It follows that the validity of these analyses hinges on the quality of these golden standards.
If a gold standard contains biases, i.e., systematic errors, it can foster bias in any downstream analysis.

@@

There is a long history of text annotation in studies analyzing political text.
While this yields lots of experiences as to how to train coders so that we get reliable hand-coded data, the procedure is expensive, protracted, and sometimes does not even get us the quality of data we need [@weber18].
While the crowd offers a solution to some of these issues [@van2021validity], the main underlying assumption that there is one correct interpretation for every input example remains untouched.
However, especially for more complex coding tasks different research fields have different interpretations of what counts as *correct* solution depending on the main goal of annotation: Do we want to create a clean benchmark data set that follows strict linguistic rules or one that reflects how statements are being interpreted by "real" readers that do not necessarily follow theoretical definitions on which elements are needed for a specified stance?
Trends of polarization have shown that people do interpret information according to their ideological position [e.g., @rekker2022understanding; @lee2021more].
*Does this mean that some are right and others are wrong? Or is there an ideological difference in the ground truth?* These questions present a fundamental challenge to the main way of working when collecting gold standard data, as we operate from the baseline assumption that disagreement among the annotators should be avoided or reduced.
Typically, when specific cases continuously cause disagreement, more instructions are added to limit interpretations (for recent innovations to improve annotation, see for example: @barbera2021automated; @struthers2020bridging; @winter2020online; @debell2013harder; @benoit2016crowd, @ying2022topics).
However, work in computational linguistics has shown that increased annotation instructions do not increase quality [@parmar2022don].
This leaves us between a rock and a hard place.
Is there a potential bias in annotators that we should account for?

In this paper, we build upon the NLP literature on disagreement -- or bias -- in annotation [e.g., see @shen2021sounds; @geva2019we; @sommerauer2020penguin; @plank2014learning] and so-called perspectivism [@cabitza2023toward; @havens2022beyond] -- i.e.\ the adoption of methods that integrate the opinions and perspectives of the human subjects involved in the knowledge representation step of the machine learning processes [@cabitza2023toward].
This literature puts forward that disagreement can occur because of differences in ideological position or political knowledge [@shen2021sounds; @alkiek2022classification; @joseph2021mis].
This allows us to test the extent to which disagreement takes place, for what type of stances, as well as gives us directions on how to deal with the diversity in conceptions and the political heterogeneity that nowadays potentially occurs in our sample of annotators.
To do so, we have fielded two high-powered pre-registered experiments (see [here](link) and [here](link)) in the Netherlands -- a low-level polarized country -- and in the U.S. -- a high-level polarized country testing the effect of ideological distance between the annotator and the political actor in the text (H1), the effect of overinterpretation based on political knowledge or ideological engagement (H2), and an offered solution of masking the political actor to mitigate the effects of ideology and knowledge (H3).[^1]
In the experiments, we vary the level of specification with which a political actor takes a stance -- a declarative sentence versus a sentence where with some knowledge on politics, the stance might be inferred -- as well as whether the political actor is shown or masked with putting `[ACTOR]` instead of the political party.
We do this for four different political issues: *Environment*, *Immigration*, *Tax Policy*, and *EU* (for the Dutch case) or *Foreign Policy* (for the American case).
This country selection does not only allow us to showcase the scope conditions of disagreement due to different levels of political heterogeneity, but also differentiates between languages.
English is not only the most dominant language for computational text analysis in the social science [@baden2022three; @dolinsky2023threegaps], crowd-coders do not need to be first language speakers, given the dominance of English in our daily lives.
This is different for Dutch, it is a language spoken by a smaller community, typically first language speakers, yet still an often-enough researched case in computational text analysis in the social science [@baden2022three; Dolinsky et al. 2023].

[^1]: The data and research compendium is published on the [main author's github page](https://github.com/MarikenvdVelden/bias-experiment) -- annonymized for the review process.

@@

Generating large data-sets has become one of the main drivers of progress in natural language understanding.
In studies of political communication, the most familiar annotation tasks involve identifying basic concepts.
This includes noting the topic of the text, the position of the actor or the tone of the text.
A recent study [@van2021validity] demonstrates that using crowd-coding platforms is a good way to collect such large data-sets.
However, having only a few workers annotate the majority of text of interest has raised concerns about data diversity and models' ability to generalize beyond the crowd-workers: In a series of experiments, @geva2019we show that often models do not generalize well to annotations from annotators that did not contribute to the training set, suggesting that annotator bias should be monitored during data-set creation.
One such potential bias, especially in times of increasing polarization [@iyengar2019origins; @gidron2019toward; @boxell2022cross], is based on ideological position of the annotator.
Given that annotators on crowd-coding platforms tend to be younger and hold more liberal values than the general public [@clifford2015samples; @huff2015these], this could potentially hamper the data diversity and generalizability of the model.
An additional reason to monitor the annotators' ideological position as a potential source of annotator bias is that a recent study in NLP showed that experiential factors influence the consistency of how political ideologies are perceived [@shen2021sounds].
Their finding challenges the "ground-truth" assumption we as researcher make that a position for example is either left-leaning or right-wing leaning.
People with different ideological backgrounds might experience that position differently.
This challenges our way of data collection.
We are interested in the effect of e.g.\ elite communication.
To study this, we allow for heterogeneous treatment effects in experimental work.
This indicates that we often do not assume that the treatment, often using text, has the same effect for different partisans.
Yet, at the same time, we forget or ignore that knowledge when creating large data-sets for our machine learning models.

The field of Natural Language Processing often works on automatically classifying texts on labels of concepts such as stance, sentiment, and political orientation.
These models are trained on data created by human annotators.
Often, this process has a final step where disagreements and differences in annotations are leveled by aggregating, averaging, or in other ways coming to a consensus on one label for one example, which is then seen as "ground truth" [@aroyo2015].
Differences from this ground truth label are seen as errors that need to be removed or sorted out.
Models then learn to predict labels for new examples based on this ground truth.
However, since several years there is some discussion on how realistic it is to have one label, especially for subjective or complex concepts and/or texts with multiple interpretations.
@aroyo2015 describes succinctly how the predominant annotation procedures for classification models run into myths such as "disagreement is bad" and "one annotation is enough".
@plank2014learning adds to this that underlying ambiguity and linguistic complexity should be considered for disagreement in annotations: not all linguistic examples are created equal.
Disagreement even occurs in seemingly objective tasks such as Part of Speech tagging [@fornaciari2021, @plank2014learning].

Another aspect to consider is that disagreement can be informative for the concept under measure - sometimes agreement can be used to validate hypotheses about how universal the perceptions of such concepts are [@sommerauer2020would].
Additional doubts on smoothing out disagreement in annotation have focussed on the lack of diversity when only annotating with one label or annotator, leading to a homogeneity especially in subjective and social tasks [@geva2019we] such as hatespeech detection or political affiliation classification.
The question then is: *Whose perspective is being recorded in these datasets, and then later in the models trained on these datasets?* Framing arbitrary representations in data as "bias" misses the political character of datasets: There is no neutral data and no apolitical standpoint from where we can call out bias.
Datasets are always "a worldview" [26] and, as such, data always remains biased." (Miceli et al., 2022, p. 5). This is key to social scientist in general, and those studying political text in particular, since several tasks of interest are intrinsically societal, with answers that differ based on the make-up of the worldview of annotators. The answer of the annotators in turn influence how machine learning examples classify new models. For instance, hate speech detection and abuse detection is one NLP task where the race and gender of annotators influences annotators and model performance [@gordon2022, @larimore2021reconsidering, @waseem2016]. Language is inherently connected to society and culture: @shenDarling analyze sentiment analysis, and find that human annotators lead to certain perspectives on sentiment being recorded, and that notably that African American English dialects are often misunderstood by such models.

Most recently, new annotation paradigms have gone one step further by asking whether we are modelling the task, or the annotator [@geva2019we].
@pavlick2019 find that for the logical coherence task Natural Language Inference, annotators have several valid interpretations that are not reflected in one ground truth label.
They call for new training paradigms that can reflect "the full range of possible human inferences"[@pavlick2019, p. 688].
Recent approaches in NLP have sought to explicitly incorporate disagreement and diversity in training data annotations.
@rottger2022two introduce the idea of an explicitly subjective annotation paradigm existing in addition to one focussed on one label and "ground truth".
Such a subjective annotation paradigm can be used for purposes where the goal is finding diverse perspectives on the task or concepts, and for models to model more accurately how humans interpret a task or text.
Additionally, "perspectivism" [@cabitza2023toward] is a paradigm and research agenda where different perspectives are explicitly incorporated in the training data, and used by models to provide more human-like classifications.
Another paradigm is "jury learning" [@gordon2022], in which machine learning models do not learn to replicate one specific ground truth, but are trained with different annotator juries to reflect the judgement of different populations.
In both approaches, demographic and other individual aspects of the annotator are explicitly mentioned and highlighted as having an influence on classification performance, but is used as an asset rather than as an error.

@troiano2021emotion [p.2] also note how for a complex concept such as the "emotion" of a text, the annotator can make several assumptions during the annotation process on what is wanted, that are all valid and may or may not be useful in different contexts: "It is possible to assess one's own emotion after reading the text, to reconstruct the affective state of the writers who produced it, to guess the reaction that they intended to elicit in the readers, and so on." @shen2021sounds find that political ideology is not an inherent concept in many texts, but rather dependent on who is asked to annotate and their perceptions and background.
Extralinguistic factors, such as annotator's own political ideology and also knowledge, influenced annotation and in turn model performance.
@thornjakobsen2022 specifically analyze a task related to stance detection, argumentative sentence detection, on these extralinguistic factors, and find an effect of gender and political leaning on annotations and also model performance.
However, to our knowledge this phenomenon has not yet been tested in a controlled experimental setting with manipulations in the texts.

To test whether people with different ideological backgrounds as well as their political knowledge might experience that position differently, challenging our ground-truth assumption in data annotation, we propose the following hypothesis:

@@

As noted by @plank2014, some linguistic examples are ambiguous, and open to multiple interpretations even for seemingly objective tasks such as part of speech tagging, where annotators have to distinguish parts of speech such as nouns and verbs.
A complex concept such as political ideology is much more likely to lead to multiple interpretations.
We call such sentences with more possible interpretations and less explicit standpoints "underspecified".
A lack of explicitness in the annotated text is one of the main causes of the disagreements in earlier literature.
@thornjakobsen2022 deduce that annotator bias comes from a process known as the affect heuristic [@slovic2007affect ]: making a decision based on the emotional response related to your own personal attitude towards the discussed topic, especially when the text is relatively ambiguous.
We therefore expect that the strength of H1a is conditional on this ambiguity:

@@

We will test these hypotheses both with a strict and a lenient interpretation of stance to illustrate whether decisions made in the research process on what constitutes a stance influence the results.
In a next step, we ask: *If these biases exists, how can we alleviate them?* There have been several previous approaches to solve biased annotations, especially where it concerns political or societal aspects.
@geva2019we introduce an approach where training set annotators are separated from annotators annotating data sets that evaluate the models, to ensure the evaluation is not simply accurate at replicating the original annotators, but can generalize to new annotators' judgements.
However, these approaches are not aimed at reducing biases during the training set annotation procedure.
Other approaches are aimed at leveraging multiple perspectives - but this is not useful when one wants one label to learn from.
For the Austrian National Elections, @ennser2018impact already demonstrated that showing party labels impact annotators' assessment of the party position.
So, one solution is masking:

@@

## Pre-registered Results

First, we test whether there is an ideological bias in interpreting stances (H1a), and if this bias increases for those that are further away from the ideological position of the political actor in the under-specified condition (H1b).
Figure \ref{fig:results-h1} demonstrates on the left-hand panel the regression coefficients and on the right-hand panel the average marginal effects of the interaction between ideological distance and under-specified sentences.
The upper-left panel of Figure \ref{fig:results-h1} demonstrates the coefficient of ideological distances for the likelihood of interpreting the stance correctly.
There is a negligible positive effect -- a coefficient of `0.002` -- that is borderline significant.
Substantially, this means that there is no effect of ideological distance for correctly interpreting the stance in either the American or Dutch case.
Hence, no ideological bias found, thus no support for our H1a.
Looking at the lower-left, and right-hand panel of Figure \ref{fig:results-h1}, we see a small but significant effect of the interaction between ideological distance and the under-specified condition for both interpretations in the American case, and only for the strict interpretation of a stance in the Dutch case.
These effects are, however, going in the other direction than hypothesized.
For those who are ideologically further away from the party, they are less likely to be wrong than those who are close too the party, as shown in the right-hand panel of Figure \ref{fig:results-h1}.
For the Dutch case, using the strict interpretation, the difference is about `20%` -- from a coefficient of `-0.6` to `-0.4`.
In the American case, the slope is even steeper, with a difference of about `40%`.
We thus find evidence for H1b.
This demonstrates that even in times of heightened polarization, ideological annotation bias can be a concern.

@@

To check the robustness of our findings, Figure \ref{fig:results-exp1} demonstrates the analyses for each issue separately.
The different colors visualize the different dependent variables.
We do not see much variation between issues *Tax*, *EU/Foreign Policy*, and *Environment*.
For those issues, we see that almost everyone interprets the sentence correct (in blue and red).
We also see that for a lenient interpretation of stances, people are quite likely to over interpret a position as a stance.
**Being correct about the stance does not when masking the political actor in both cases**.
Yet, the chance of being correct decreases statistically significantly when the sentence is underspecified.
The same holds for overinterpreting for the lenient interpretation, but the opposite is true for the strict interpretation; there overinterpretation is more likely with underspecified sentences.
Looking at *Immigration*, we see a different pattern.
We see that masking does not increase the likelihood of being correct, but does increase the likelihood of overinterpretation regardless of how one defines a stance.
Underspecified sentences are less likely to be correctly identified and more likely to be overinterpreted regardless of the definition of a stance.
So, while there are some differences in effect sizes between the issues, the overall findings are not driven by a single issue.

@@

## Recommendations

TBA

\newpage

# References

>>>>

---
output: 
  stevetemplates::article:
    fig_caption: true
    extra_dependencies: ["float"]
bibliography: references.bib
urlcolor: black
linkcolor: black
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage[para,online,flushleft]{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{hyperref}
   - \usepackage{array}   
   - \usepackage{caption}
   - \usepackage{graphicx}
   - \usepackage{epstopdf}
   - \usepackage{siunitx}
   - \usepackage{hhline}
   - \usepackage{calc}
   - \usepackage{tabularx}
   - \usepackage{fontawesome}
   - \usepackage{amsthm}
   - \newtheorem{hypo}{Hypothesis}
   - \usepackage{setspace}
biblio-style: apsr
title: "Whose Truth is it Anyway? An Experiment on Annotation Bias in Times of Factual Opinion Polarization"
author:
- name: Mariken A.C.G. van der Velden
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
- name: Myrthe Reuver
- affiliation: Computational Linguistics & Text Mining Lab, Vrije Universiteit Amsterdam
- name: Wouter van Atteveldt
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
- name: Antse Fokkens
- affiliation: Computational Linguistics & Text Mining Lab, Vrije Universiteit Amsterdam
- name: Felicia Loecherbach
- affiliation: Center for Social Media & Politics, New York University
- name: Kasper Welbers
- affiliation: Dep. of Communciation Science, Vrije Universiteit Amsterdam
thanks: "Corresponding author: MACGvdV, Replication files are available on the author's Github account (https://github.com/MarikenvdVelden/bias-experiment); Author contributions: a) designed the study: MACGvdV, MR, WvA, AF, FL, & KW; b) conducted the study: MACGvdV, FL & MR; c) data cleaning & analysis: MACGvdV; d) writing of the paper: MACGvdV, MR, WvA, AF, FL, & KW"
anonymous: TRUE
abstract: "The behavior of citizens and the study of them by social scientists rely heavily on key information. Human annotation represents a crucial instrument in the toolkit of the social sciences. To address concerns regarding the validity and reliability of annotation tasks, we establish strict standards. Due to the democratization of data and the advances in NLP more data can be analyzed or classified, these standards are more important than ever: An algorithm trained on biased data will reproduce and often exacerbate bias. Our tools to create valid and reliable annotation data does currently not allow for dealing with different perspectives of annotators. In two pre-registered experiments in the United States and the Netherlands, we show that personal characteristics of annotators, like political ideology or knowledge, interfere with annotators judgement of political stances. Our results show that to improve annotated data for automated text analyses, and for stance detection models in particular, we need to critically evaluate how we create our gold standards."
keywords: "Experiment, Annotation Bias, Ideology, Measurinig Political Position, Text-as-Data, Political Knowledge"
geometry: margin=1in
mainfont: cochineal
fontsize: 12pt
params:
  anonymous: ""
endnote: no
pandocparas: TRUE
sansitup: FALSE
editor_options: 
  markdown: 
    wrap: sentence
---

@@

\doublespacing
# Introduction
The behavior of citizens and the study of them by social scientists rely heavily on key information.
Classical theories of political communication, such as agenda setting or framing [e.g. @vanaelst2016political; @lecheler2019news], formulate that political information drives opinion formation and participation in politics -- from voting to protests [for overviews hereof across countries, see @pfetsch2013comparing].
The democratization of data and advent of computational social science has paved the way for new possibilities of gathering and analyzing textual data [for a recent overview, see @vanatteveldt2022computational]. 
In particular, advances in Natural Language Processing (NLP) have made it possible to automatically analyze large quantities of data using machine learning [e.g., see @grimmer2022text; @wei2023overview].
To determine the validity of such large scale computational analyses, we rely on "gold standard" data, created by human annotators.
It follows that the validity of these analyses hinges on the quality of these golden standards.
If a gold standard contains biases, i.e., systematic errors, it can foster bias in any downstream analysis.

@@

There is a long history of text annotation in studies analyzing political text.
While this yields lots of experiences as to how to train coders so that we get reliable hand-coded data, the procedure is expensive, protracted, and sometimes does not even get us the quality of data we need [@weber18].
While the crowd offers a solution to some of these issues [@van2021validity], the main underlying assumption that there is one correct interpretation for every input example remains untouched.
However, especially for more complex coding tasks different research fields have different interpretations of what counts as *correct* solution depending on the main goal of annotation.
Do we want to create a clean benchmark data set that follows strict linguistic rules or one that reflects how statements are being interpreted by "real" readers that do not necessarily follow theoretical definitions on which elements are needed for a specified stance?
Trends of polarization have shown that people do interpret information according to their ideological position [e.g., @rekker2022understanding; @lee2021more].
*Does this mean that some are right and others are wrong? Or is there an ideological difference in the ground truth?* 
These questions present a fundamental challenge to the main way of working when collecting gold standard data, as we operate from the baseline assumption that disagreement among the annotators should be avoided or reduced.
Typically, when specific cases continuously cause disagreement, more instructions are added to limit interpretations (for recent innovations to improve annotation, see for example: @barbera2021automated; @struthers2020bridging; @winter2020online; @debell2013harder; @benoit2016crowd, @ying2022topics).
However, work in computational linguistics has shown that increased annotation instructions do not increase quality [@parmar2022don].
This leaves us between a rock and a hard place.
Is there a potential bias in annotators that we should account for?

In this paper, we build upon the NLP literature on disagreement -- or bias -- in annotation [e.g., see @shen2021sounds; @geva2019we; @sommerauer2020penguin; @plank2014learning] and so-called perspectivism [@cabitza2023toward; @havens2022beyond] -- i.e. the adoption of methods that integrate the opinions and perspectives of the human subjects involved in the knowledge representation step of the machine learning processes [@cabitza2023toward].
This literature puts forward that disagreement can occur because of differences in ideological position or political knowledge [@shen2021sounds; @alkiek2022classification; @joseph2021mis].
This allows us to test the extent to which disagreement takes place, for what type of stances, as well as gives us directions on how to deal with the diversity in conceptions and the political heterogeneity that nowadays potentially occurs in our sample of annotators.
To do so, we have fielded two high-powered pre-registered experiments (see [here](link) and [here](link)) in the Netherlands -- a low-level polarized country -- and in the U.S. -- a high-level polarized country testing the effect of ideological distance between the annotator and the political actor in the text (H1), the effect of overinterpretation based on political knowledge or ideological engagement (H2), and an offered solution of masking the political actor to mitigate the effects of ideology and knowledge (H3).^[The data and research compendium is published on the [[main author's github page]](https://github.com/MarikenvdVelden/bias-experiment) -- annonymized for the review process.]
In the experiments, we vary the level of specification with which a political actor takes a stance -- a declarative sentence versus a sentence where with some knowledge on politics, the stance might be inferred -- as well as whether the political actor is shown or masked with putting `[ACTOR]` instead of the political party.
We do this for four different political issues: *Environment*, *Immigration*, *Tax Policy*, and *EU* (for the Dutch case) or *Foreign Policy* (for the American case).
This country selection does not only allow us to showcase the scope conditions of disagreement due to different levels of political heterogeneity, but also differentiates between languages.
English is not only the most dominant language for computational text analysis in the social science [@baden2022three; @dolinsky2023threegaps], crowd-coders do not need to be first language speakers, given the dominance of English in our daily lives.
This is different for Dutch, it is a language spoken by a smaller community, typically first language speakers, yet still an often-enough researched case in computational text analysis in the social science [@baden2022three; @dolinsky2023threegaps].

@@

Generating large data-sets has become one of the main drivers of progress in natural language understanding.
In studies of political communication, the most familiar annotation tasks involve identifying theoretical concepts.
This includes noting the topic of the text, the position of the actor or the tone of the text.
Crowdcoding seems suitable for this task [@van2021validity].
Yet, having only a few workers annotate the majority of text of interest has raised concerns about data diversity and models' ability to generalize beyond the crowd-workers.
In a series of experiments, @geva2019we show that often models do not generalize well to annotations from annotators that did not contribute to the training set, suggesting that annotator bias should be monitored during data-set creation.
One such potential bias, especially in times of increasing polarization [@iyengar2019origins; @gidron2019toward; @boxell2022cross], is based on ideological position of the annotator.
Given that annotators on crowd-coding platforms tend to be younger and hold more liberal values than the general public [@clifford2015samples; @huff2015these], this could potentially hamper the data diversity and generalizability of the model.
An additional reason to monitor the annotators' ideological position as a potential source of annotator bias is that a recent study in NLP showed that experiential factors influence the consistency of how political ideologies are perceived [@shen2021sounds].
Their finding challenges the "ground-truth" assumption we as researcher make that a position for example is either left-leaning or right-wing leaning.
People with different ideological backgrounds might experience that position differently.
This challenges our way of data collection.
We are interested in the effect of e.g. elite communication.
To study this, we allow for heterogeneous treatment effects in experimental work.
This indicates that we often do not assume that the treatment, often using text, has the same effect for different partisans.
Yet, at the same time, we forget or ignore that knowledge when creating large data-sets for our machine learning models.

The field of Natural Language Processing often works on automatically classifying texts on labels of concepts such as stance, sentiment, and political orientation.
These models are trained on data created by human annotators.
Often, this process has a final step where disagreements and differences in annotations are leveled by aggregating, averaging, or in other ways coming to a consensus on one label for one example, which is then seen as "ground truth" [@aroyo2015].
Differences from this ground truth label are seen as errors that need to be removed or sorted out.
Models then learn to predict labels for new examples based on this ground truth.
However, since several years there is some discussion on how realistic it is to have one label, especially for subjective or complex concepts and/or texts with multiple interpretations.
@aroyo2015 describes succinctly how the predominant annotation procedures for classification models run into myths such as "disagreement is bad" and "one annotation is enough".
@plank2014learning adds to this that underlying ambiguity and linguistic complexity should be considered for disagreement in annotations: not all linguistic examples are created equal.
Disagreement even occurs in seemingly objective tasks such as Part of Speech tagging [@fornaciari2021; @plank2014learning].

Another aspect to consider is that disagreement can be informative for the concept under measure - sometimes agreement can be used to validate hypotheses about how universal the perceptions of such concepts are [@sommerauer2020would].
Additional doubts on smoothing out disagreement in annotation have focussed on the lack of diversity when only annotating with one label or annotator, leading to a homogeneity especially in subjective and social tasks [@geva2019we] such as hatespeech detection or political affiliation classification.
The question then is: *Whose perspective is being recorded in these datasets, and then later in the models trained on these datasets?* Framing arbitrary representations in data as "bias" misses the political character of datasets: There is no neutral data and no apolitical standpoint from where we can call out bias.
Datasets are always "a worldview" and, as such, data always remains biased" [@miceli2022studying p.5]. 
This is key to social scientist in general, and those studying political text in particular, since several tasks of interest are intrinsically societal, with answers that differ based on the make-up of the worldview of annotators. 
The answer of the annotators in turn influence how machine learning examples classify new models. 
For instance, hate speech and abuse detection are NLP tasks where race and gender of annotators influences both annotators and model performance [@gordon2022; @larimore2021reconsidering; @waseem2016]. 
Language is inherently connected to society and culture: @shenDarling analyze sentiment analysis, and find that human annotators lead to certain perspectives on sentiment being recorded; notably African American English dialects are often misunderstood by such models.

Most recently, new annotation paradigms have gone one step further by asking whether we are modelling the task, or the annotator [@geva2019we].
@pavlick2019 find that for the logical coherence task Natural Language Inference, annotators have several valid interpretations that are not reflected in one ground truth label.
They call for new training paradigms that can reflect "the full range of possible human inferences"[@pavlick2019, p.688].
Recent approaches in NLP have sought to explicitly incorporate disagreement and diversity in training data annotations.
@rottger2022two introduce the idea of an explicitly subjective annotation paradigm existing in addition to one focussed on one label and "ground truth".
Such a subjective annotation paradigm can be used for purposes where the goal is finding diverse perspectives on the task or concepts, and for models to model more accurately how humans interpret a task or text.
Additionally, "perspectivism" [@cabitza2023toward] is a paradigm and research agenda where different perspectives are explicitly incorporated in the training data, and used by models to provide more human-like classifications.
Another paradigm is "jury learning" [@gordon2022], in which machine learning models do not learn to replicate one specific ground truth, but are trained with different annotator juries to reflect the judgement of different populations.
In both approaches, demographic and other individual aspects of the annotator are explicitly mentioned and highlighted as having an influence on classification performance, but is used as an asset rather than as an error.

@troiano2021emotion [p.2] also note how for a complex concept such as the "emotion" of a text, the annotator can make several assumptions during the annotation process on what is wanted, that are all valid and may or may not be useful in different contexts: "It is possible to assess one's own emotion after reading the text, to reconstruct the affective state of the writers who produced it, to guess the reaction that they intended to elicit in the readers, and so on." @shen2021sounds find that political ideology is not an inherent concept in many texts, but rather dependent on who is asked to annotate and their perceptions and background.
Extralinguistic factors, such as annotator's own political ideology and also knowledge, influenced annotation and in turn model performance.
@thornjakobsen2022 specifically analyze a task related to stance detection, argumentative sentence detection, on these extralinguistic factors, and find an effect of gender and political leaning on annotations and also model performance.
However, to our knowledge this phenomenon has not yet been tested in a controlled experimental setting with manipulations in the texts.
To test whether people with different ideological backgrounds experience a political position differently, challenging our ground-truth assumption in data annotation, we propose the following hypothesis:

@@

As noted by @plank2014, some linguistic examples are ambiguous, and open to multiple interpretations even for seemingly objective tasks such as part of speech tagging, where annotators have to distinguish parts of speech such as nouns and verbs.
A complex concept such as political ideology is much more likely to lead to multiple interpretations.
We call such sentences with more possible interpretations and less explicit standpoints "underspecified".
A lack of explicitness in the annotated text is one of the main causes of the disagreements in earlier literature.
@thornjakobsen2022 deduce that annotator bias comes from a process known as the affect heuristic [@slovic2007affect]: Making a decision based on the emotional response related to your own personal attitude towards the discussed topic, especially when the text is relatively ambiguous.
We therefore expect that the strength of H1a is conditional on this ambiguity:

@@

In a next step, we ask: *If these biases exists, how can we alleviate them?* There have been several previous approaches to solve biased annotations, especially where it concerns political or societal aspects.
@geva2019we introduce an approach where training set annotators are separated from annotators annotating data sets that evaluate the models, to ensure the evaluation is not simply accurate at replicating the original annotators, but can generalize to new annotators' judgements.
However, these approaches are not aimed at reducing biases during the training set annotation procedure.
Other approaches are aimed at leveraging multiple perspectives - but this is not useful when one wants one label to learn from.
For the Austrian National Elections, @ennser2018impact already demonstrated that showing party labels impact annotators' assessment of the party position.
So, one solution is masking. 
We therefore test whether masking reduces potential differentiation incited by ideoloigcal postion or political knowledge with the following hypotheses:

@@

## Pre-registered Results
First, we test whether there is an ideological bias in interpreting stances (H1a), and if this bias increases for those that are further away from the ideological position of the political actor in the under-specified condition (H1b).
Figure \ref{fig:results-h1} demonstrates on the left-hand panel the regression coefficients and on the right-hand panel the average marginal effects of the interaction between ideological distance and under-specified sentences.
The upper-left panel of Figure \ref{fig:results-h1} demonstrates the coefficient of ideological distances for the likelihood of interpreting the stance correctly.
There is a negligible positive effect -- a coefficient of `0.002` -- that is borderline significant.
Substantially, this means that there is no effect of ideological distance for correctly interpreting the stance in either the American or Dutch case.
Hence, no ideological bias found, thus no support for our H1a.
Looking at the lower-left, and right-hand panel of Figure \ref{fig:results-h1}, we see a small but significant effect of the interaction between ideological distance and the under-specified condition for both interpretations in the American case, and only for the strict interpretation of a stance in the Dutch case.
These effects are, however, going in the other direction than hypothesized.
For those who are ideologically further away from the party, they are less likely to be wrong than those who are close too the party, as shown in the right-hand panel of Figure \ref{fig:results-h1}.
For the Dutch case, using the strict interpretation, the difference is about `20%` -- from a coefficient of `-0.6` to `-0.4`.
In the American case, the slope is even steeper, with a difference of about `40%`.
We thus find evidence for H1b.
This demonstrates that even in times of heightened polarization, ideological annotation bias can be a concern.

@@

To check the robustness of our findings, Figure \ref{fig:results-exp1} demonstrates the analyses for each issue separately.
The different colors visualize the different dependent variables.
We do not see much variation between issues *Tax*, *EU/Foreign Policy*, and *Environment*.
For those issues, we see that almost everyone interprets the sentence correct (in blue and red).
We also see that for a lenient interpretation of stances, people are quite likely to over interpret a position as a stance.
Being correct about the stance does not decrease when masking the political actor in both cases.
Yet, the chance of being correct decreases statistically significantly when the sentence is underspecified.
The same holds for overinterpreting for the lenient interpretation, but the opposite is true for the strict interpretation; there overinterpretation is more likely with underspecified sentences.
Looking at *Immigration*, we see a different pattern.
We see that masking does not increase the likelihood of being correct, but does increase the likelihood of overinterpretation regardless of how one defines a stance.
Underspecified sentences are less likely to be correctly identified and more likely to be overinterpreted regardless of the definition of a stance.
So, while there are some differences in effect sizes between the issues, the overall findings are not driven by a single issue.

@@

\newpage
\singlespacing
# References
